---
title: "MCMCglmm"
output:
  bookdown::html_document2:
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
pkgdown:
  as_is: true 
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is a modified version of a course I gave about generalised linear mixed models (GLMM). Special emphasis is placed on understanding the underlying structure of a GLMM in order to show that slight modifications of this structure can produce a wide range of models.  These include familiar models like regression, but also models with intimidating names: animal models, threshold models, meta-analysis, Rasch models and random regression... The primary aim of this course is to show that these models are only daunting by name. The secondary aim is to show how these models can be fitted in a Bayesian framework using Markov chain Monte Carlo (MCMC) methods in the R package MCMCglmm.

```{r }
library(MCMCglmm)
```
 
```{r child = '_2.Bayes_and_MCMC.Rmd'}
```

```{r child = '_3.GLMM.Rmd'}
```


```{r child = '_4.Categorical_R.Rmd'}
```


```{r child = '_5.Continuous_R.Rmd'}
```


```{r child = '_6.Multi_Response.Rmd'}
```


```{r child = '_7.Technical_Details.Rmd'}
```


```{r child = '_8.Parameter_Expansion.Rmd'}
```

```{r child = '_9.Path_Analysis.Rmd'}
```
