# Structured Random Effects {#pedigree}


In previous chapters we introduced variance functions that allow interactions between a categorical predictor (which usually has many levels and would normally be fitted as random in the absence of a interaction) and some other categorical (Chapters \@ref(cat-int) and \@ref(multi)) or continuous predictor (Chapter \@ref(cont-int)). For example, in Chapter \@ref(cat-int) we fitted a model with $\texttt{us(virus):line}$ in the random formula. We could visualise this as setting up `r 3*nlevels(sigma_small$line)` random effects for the three types of virus (France, Greece and Spain) across `r nlevels(sigma_small$line)` genetically distinct lines of fly: 


$$\begin{array}{c|rrrrrc}
&{\color{red}{\texttt{101A}}}&{\color{red}{\texttt{109A}}}&{\color{red}{\texttt{129A}}}&{\color{red}{\texttt{142A}}}&{\color{red}{\texttt{153A}}}&\dots\\
\hline
{\color{blue}{\texttt{France}}}&{\color{blue}{\texttt{France}}}.{\color{red}{\texttt{101A}}}&{\color{blue}{\texttt{France}}}.{\color{red}{\texttt{109A}}}&{\color{blue}{\texttt{France}}}.{\color{red}{\texttt{129A}}}&{\color{blue}{\texttt{France}}}.{\color{red}{\texttt{142A}}}&{\color{blue}{\texttt{France}}}.{\color{red}{\texttt{153A}}}&\dots\\
{\color{blue}{\texttt{Spain}}}&{\color{blue}{\texttt{Spain}}}.{\color{red}{\texttt{101A}}}&{\color{blue}{\texttt{Spain}}}.{\color{red}{\texttt{109A}}}&{\color{blue}{\texttt{Spain}}}.{\color{red}{\texttt{129A}}}&{\color{blue}{\texttt{Spain}}}.{\color{red}{\texttt{142A}}}&{\color{blue}{\texttt{Spain}}}.{\color{red}{\texttt{153A}}}&\dots\\
{\color{blue}{\texttt{Greece}}}&{\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{101A}}}&{\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{109A}}}&{\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{129A}}}&{\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{142A}}}&{\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{153A}}}&\dots\\
\end{array}$$

The $\texttt{us}$ structure allows the line effects for each virus to have different variances, but it also allows the effects for the same line to be correlated across viruses: the effects ${\color{blue}{\texttt{France}}}.{\color{red}{\texttt{129A}}}$ and ${\color{blue}{\texttt{Greece}}}.{\color{red}{\texttt{129A}}}$ are allowed to be correlated. However, effects from different lines are assumed to be independent, both within and between virsus: the effects ${\color{blue}{\texttt{France}}}.{\color{red}{\texttt{129A}}}$ and ${\color{blue}{\texttt{France}}}.{\color{red}{\texttt{142A}}}$ are assumed independent with zero correlation because they are associated with different lines.

In some cases we would like to allow correlations between the effects associated with different lines. For example, lines 129A and 142A may be genetically very similar and so we may expect them to respond to viruses in a similar way. If the covariance structure of a set of random effects is known up to proportionality, $\texttt{MCMCglmm}$ allows this to be specified and the constant or low-dimensional matrix of constants is estimated. We will start with the covariance structures that arise because individuals or species have shared ancestry.  In the context of individuals, these models are known as 'animal' models and shared ancestry is represented by a pedigree. In the context of species, these models are known as phylogenetic mixed models and shared ancestry is represented by a phylogeny. The two models are almost identical, and are relatively minor modifications to the basic mixed model. 


## Pedigrees

$\texttt{MCMCglmm}$ handles pedigrees stored in 3-column tabular form, with each row representing a single individual. The first column should contain the unique identifier of the individual, and columns 2 and 3 should be the unique identifiers of the individual's parents. Parents must appear before their offspring in the table - the function $\texttt{orderPed}$ from the library $\texttt{MasterBayes}$ will reorder the pedigree of required. I usually have the dam (mother) in the first column and the sire (father) in the third. I prefer the words dam and sire because if I subscript things with $m$ and $f$ I can never remember whether I mean male and female, or mother and father. In hermaphrodite systems the same individual may appear in both columns, even within the same row if an individual was produced through selfing. This is not a problem, but $\texttt{MCMCglmm}$ will issue a warning in case hermaphrodites are not present and a data entry mistake has been made. Impossible pedigrees (for example individual's that give birth to their own mother) are a problem and $\texttt{MCMCglmm}$ will issue an error, hopefully with an appropriate message, when impossibilities are detected.

If the parent(s) of an individual are unknown then a missing value (`NA`) should be assigned in the relevant column. All individuals appearing as dams or sires need to have their own record, even if both of their parents are unknown. The function $\texttt{insertPed}$ will return a pedigree table with these individuals inserted if they are missing. Often the number of individuals in a pedigree will be greater than the number of individuals for which phenotypic data exist. $\texttt{MCMCglmm}$ can handle this, as long as all the individuals appearing in the data frame passed to `data` also appear in the pedigree. However, it will be more efficient to prune out individuals from the pedigree that lack phenotypic data and which are not required in order to determine the relatedness of two or more phenotyped individuals (e.g. the unphenotyped parents of two siblings) - the function $\texttt{prunePed}$ will do this [^prune].

[^prune]: If you pass a pedigree to $\texttt{prunePed}$ with the id's of phenotyped individuals specified in the argument $\texttt{keep}$ a pruned pedigree is returned. Pruning the pedigree has no impact on the posterior distribution but it can improve mixing and speed.  In earlier versions of MCMCglmm (<3.0) the default argument to $\texttt{prunePed}$ was $\texttt{make.base=FALSE}$ but this did not result in the optimal amount of mixing and was only retained for back compatibility.  

To illustrate, we can load a pedigree for a population of blue tits and display the pedigree for the nuclear family that has individuals $\texttt{R187920}$ and $\texttt{R187921}$ as parents:

```{r }
data(BTped)
Nped <- BTped[which(apply(BTped, 1, function(x) {any(x == "R187920" | x == "R187921")})), ]
Nped
```

Both parents form part of what is known as the base population - they are assumed to be outbred and unrelated to anybody else in the pedigree.  A pedigree can also be expressed in terms of a relatedness matrix ${\bf A}$. This matrix is symmetric, square, and has dimensions equal to the number of individuals in the pedigree. Simple, but perhaps slow, recursive methods exist for calculating ${\bf A}$:

```{r }
Aped <- 2 * kinship2::kinship(Nped[, 1], Nped[, 2], Nped[, 3])
Aped
```

In 'animal' models, each individual is associated with a unique random effect (often called a 'breeding value') and their covariance structure is assumed to proportional to ${\bf A}$ rather than an identity matrix. This non-identity covariance structure allows the associated variance to be estimated even when there is only one observation per individual.  Although ${\bf A}$ has quite a straightforward interpretation, $\texttt{MCMCglmm}$ actually requires the inverse of ${\bf A}$ which can be obtained using the function $\texttt{inverseA}$. We can create this for the full pedigree which consists of `r nlevels(BTped$dam)` nuclear families with two parents and between `r min(table(BTped$sire))` and `r max(table(BTped$sire))` of their offspring offspring. 


```{r }
Ainv<-inverseA(BTped)$Ainv
```

The data set $\texttt{BTdata}$ contains information collected on their offspring: the length of one of their leg bones ($\texttt{tarsus}$) together with their identifier ($\texttt{animal}$) their sex ($\texttt{sex}$) and the nest they were raised in ($\texttt{fosternest}$). Note that roughly half the chicks in each nest were switched into another nest soon after hatching and so the identity of the nuclear family and $\texttt{fosternest}$ are not completely confounded. We can fit a model which includes both $\texttt{fosternest}$ nest effects and also $\texttt{animal}$ effects that are correlated through a pedigree:

```{r animal}
data(BTdata)
prior.animal<-list(R=IW(1, 0.002), G=F(1, 1000))
m.animal <- MCMCglmm(tarsus ~ sex, random = ~fosternest+animal, data = BTdata, ginv=list(animal=Ainv), prior=prior.animal, longer=10)
```

The $\texttt{ginv}$ argument takes a list of inverse covariance structures (just $\texttt{Ainv}$ in this example) where the names of the list elements state the name of the effects the covariance structure is associated with (just $\texttt{animal}$ in this example). 


```{r }
summary(m.animal)
```

Often, the variances of the random effects are expressed as proportions of the total variance:

```{r }
plot(m.animal$VCV/rowSums(m.animal$VCV))
```

and can see that the while the nest a chick is raised in ($\texttt{fosternest}$) explains relatively little variation, the genetic variance (the variance in $\texttt{animal}$ effects) explains around half the total. It is also apparent from the MCMC traces that despite the chain being run for 10 times longer than the default ($\texttt{longer=10}$) there is still a bit of autocorrelation. In general, animal models mix less well than most models, in part because of strong posterior correlations between the genetic and residual variance:

```{r }
plot(c(m.animal$VCV[,"animal"])~c(m.animal$VCV[,"units"]), ylab=expression(sigma[animal]^2), xlab=expression(sigma[units]^2))
```

This correlation exists because each offspring receives a unique jumble of it's parents genome, the effect of which (the Mendelian sampling deviation) is confounded with the effect of the unique environmental factors the offspring experiences (the residual). This can be observed in the relatedness matrix $\texttt{Aped}$ where the diagonal elements are one yet the off-diagonal elements are at best a half.  In this particular example, the relatedness structure of phenotyped individuals is very simple: either individuals share a mother (or father) and are related by a half since they are full-siblings or they are completely unrelated. Consequently, an equivalent model is to fit nuclear family as an effect (by fitting mother ($\texttt{dam}$) effects) rather than pedigree ($\texttt{animal}$) effects:


```{r sib}
prior.sib<-list(R=IW(1, 0.002), G=F(1, 1000))
m.sib <- MCMCglmm(tarsus ~ sex, random = ~fosternest+dam, data = BTdata, prior=prior.sib)
```
In this model, the $\texttt{units}$ variance is the variance due to environmental effects and the genetic effects unique to an individual. Under the assumptions of the animal model, the variance of the unique genetic effects is assumed to be half the genetic variance. Under the assumptions of the animal model, the $\texttt{dam}$ variance should be half the genetic variance since it is the variance due to the 50% of the genome full siblings share. Consequently, $2\sigma^2_{dam}$ from `m.sib` should be equivalent to the $\texttt{animal}$ variance from `m.animal`, and  $\sigma^2_{units}-\sigma^2_{dam}$ from `m.sib` should be equivalent to the $\texttt{units}$ variance from `m.animal`: 

```{r echo=FALSE}
sibV<-m.sib$VCV
sibV[,"units"]<-sibV[,"units"]-sibV[,"dam"]
sibV[,"dam"]<-sibV[,"dam"]*2
colnames(sibV)<-colnames(m.animal$VCV)
attr(sibV, "mcpar")<-attr(m.animal$VCV, "mcpar")
plot(mcmc.list(sibV, m.animal$VCV), density=FALSE)
```

Slight discrepancies exist because the prior specifications are not equivalent.  

## Phylogenies

Phylogenies can also be expressed in tabular form, although only two columns are required because each species only has a single parent. However, whereas in pedigrees a single generation elapses between a parent and an offspring, a variable number of generations or a variable amount of time may elapse between parents and their daughter species. 


In general however, phylogenies are not expressed in this form presumably because it is hard to traverse phylogenies (and pedigrees) backwards in time when they are stored this way. For phylogenetic mixed models we generally only need to traverse phylogenies forward in time (if at all) but I have stuck with convention and used the `phylo` class from the `ape` package to store phylogenies. As with pedigrees, all species appearing in the data frame passed to `data` need to appear in the phylogeny. Typically, this will only include species at the tips of the phylogeny and so the measured species should appear in the `tip.label` element of the phylo object. An error message will be issued if this is not the case. Data may also exist for ancestral species, or even for species present at the tips but measured many generations before. It is possible to include these data as long as the phylogeny has labelled internal nodes. If nodes are unlabelled then $\texttt{MCMCglmm}$ names them internally using the default arguments of `makeNodeLabel` from `ape`.

To illustrate, lets take the phylogeny of bird families included in the `ape` package, and plotted in Figure \@ref(fig:bird.families). 


``` {r label=bird.families, echo=FALSE, include=TRUE, fig.cap="A phylogeny of bird families from \\citet{Sibley.1990} The families in red are the Tits (Paridae), Treecreepers (Certhiidae),  Cranes (Gruidae)  and the Ostriches (Struthionidae) from top to bottom. Blue tits are in the Paridae, and the word pedigree comes from the french for crane's foot.", fig.width=7, fig.height=7}
bird.families <- makeNodeLabel(bird.families)
some.families <- c("Certhiidae", "Paridae", "Gruidae", "Struthionidae")
plot(bird.families, cex=0.5, tip.col=c("black","red")[bird.families$tip.label%in%some.families+1])
```


We can, if we wish, extract the phylogeny in tabular form. For the Paridae (Tits), Certhiidae (Treecreepers), Gruidae (Cranes) and the Struthionidae (Ostriches) this looks like:


``` {r echo=TRUE}
data("bird.families")
Nphylo <- drop.tip(bird.families, setdiff(bird.families$tip.label, some.families))
INphylo <- inverseA(Nphylo)
INphylo$pedigree
```



The full phylogeny, with these families and their connecting notes displayed, is shown in Figure \@ref(fig:bird.families). You will notice that $\texttt{Node1}$ - the root - does not appear in the phylogeny in tabular form. This is because the root is equivalent to the base population in a pedigree analysis, an issue which we will come back to later. Another piece of information that seems to be lacking in the tabular form is the branch length information. Branch lengths are equivalent to inbreeding coefficients in a pedigree. As with pedigrees the inbreeding coefficients are calculated by `inverseA`:

``` {r echo=TRUE}
INphylo$inbreeding
```

You will notice that the Struthionidae have an inbreeding coefficient of 1 because we used the default `scale=TRUE` in the call to `inverseA`. Only ultrametric trees can be scaled in $\texttt{MCMCglmm}$ and in this case the sum of the inbreeding coefficients connecting the root to a terminal node is one. To take the Paridae as an example:

``` {r echo=TRUE}
sum(INphylo$inbreeding[which(INphylo$pedigree[, 1]%in%c("Paridae","Node122", "Node58"))])
```

The inbreeding coefficients for the members of the blue tit nuclear family are of course all zero:

``` {r echo=TRUE}
inverseA(Nped)$inbreeding
```

## The animal model and the phylogenetic mixed model

The structure of pedigrees and phylogenies can be expressed in terms of the relatedness matrix ${\bf A}$. This matrix is symmetric, square, and has dimensions equal to the number of individuals in the pedigree (or the number of taxa in the phylogeny). For pedigrees, element $A_{i,j}$ is twice the probability that an allele drawn from individual $i$ is identical by descent to an allele in individual $j$. For phylogenies, element $A_{i,j}$ is the amount of time that elapsed (since the common ancestor of all sampled taxa) before the speciation event that resulted in taxa $i$ and $j$. Simple, but perhaps slow, recursive methods exist for calculating ${\bf A}$ in both cases:

``` {r echo=TRUE}
Aped <- 2 * kinship2::kinship(Nped[, 1], Nped[, 2], Nped[, 3])
Aped
```

``` {r echo=TRUE}
Aphylo <- vcv.phylo(Nphylo, cor = T)
Aphylo
```

Note that specifying `cor=T` is equivalent to scaling the tree as we did in the argument to `inverseA`.

In fact, all of the mixed models we fitted in earlier sections also used an ${\bf A}$ matrix, but in those cases the matrix was an identity matrix (i.e. ${\bf A}={\bf I}$) and we didn't have to worry about it. Let's reconsider the Blue tit model `m3a.1` from Section \@ref(cat-int) where we were interested in estimating `sex` effects for `tarsus` length together with the amount of variance explained by genetic mother (`dam`) and foster mother (`fosternest`):

``` {r echo=TRUE, cache=TRUE}
data(BTdata)
m3a.1 <- MCMCglmm(tarsus ~ sex, random = ~dam + fosternest, data = BTdata, verbose = FALSE)
```

All individuals that contributed to that analysis are from a single generation and appear in `BTped` together with their parents. However, individuals in the parental generation do not have tarsus length measurements so they do not have their own records in `BTdata`.

The model can be expressed as:

$${\bf y}  = {\bf X}{\boldsymbol{\mathbf{\beta}}}+{\bf Z}_{1}{\bf u}_{1}+{\bf Z}_{2}{\bf u}_{2}+{\bf e}$$

where the design matrices contain information relating each individual to a `sex` (${\bf X}$) a `dam` (${\bf Z}_{1}$) and a `fosternest`(${\bf Z}_{2}$). The associated parameter vectors (${\boldsymbol{\mathbf{\beta}}}$, ${\bf u}_{1}$ and ${\bf u}_{2}$) are the effects of each `sex`, `mother` and `fosternest` on `tarsus` length, and ${\bf e}$ is the vector of residuals.

In the model, the $u$'s are treated as random so we estimate their variance instead of fixing it in the prior at some (large) value, as we did with the $\beta$'s. We can be a little more explicit about what this means:

$${\bf u}_{1}  \sim N({\bf 0},\ {\bf I}\sigma^{2}_{1})$$

where $\sim$ stands for 'is distributed as' and $N$ a (multivariate) normal distribution. The distribution has two sets of parameters; a vector of means and a covariance matrix. We assume the random effects are deviations around the fixed effect part of the model and so they have a prior expectation of zero. The (co)variance matrix of the random
effects is ${\bf I}\sigma^{2}_{1}$ where $\sigma^{2}_{1}$ is the variance component to be estimated. The use of the identity matrix makes two things explicit. First, because all off-diagonal elements of an identity matrix are zero we are assuming that all `dam` effects are independent (no covariance exists between any two `dam` effects). Second, all diagonal elements of an identity matrix are 1 implying that the range of possible values the dam effect *could* take is equivalent for every `dam`, this range being governed by the magnitude of the variance component.

Since `dam`'s have very little interaction with the subset of offspring that were moved to a `fosternest`, we may be willing to assume that any similarity that exists between the tarsus lengths of this susbset and the subset that remained at home must be due to genetic effects. Although not strictly true we can assume that individuals that shared the same dam also shared the same sire, and so share around 50% of their genes.

While the numerical values are harder to interpret the patterns of zeros (represented as dots) and non-zeros in inverse covariance matrices are revealing: a zero indicates that two variables are *conditionally* independent. With Mendelian genetics you receive a random half of your dad's genome and a random half of your mum's genome. Consequently, once you condition on a parent's breeding value, siblings' breeding values are independent since they receive a random half of their parent's genome. There are no individuals that could make offspring conditionally independent of their parents since they receive their genetic material directly from. Perhaps more surprisingly, we also have a non-zero element between the mum and dad, despite them being related. The reason for this is that information on a spouse's genome provides information about the focal individual's genome if they have offspring together. To see this, imagine all offspring are heterozygous at a site (carry an $a$ allele and a $A$ allele). If we knew the spouse was $AA$ then the focal individual must either be $aa$ or $Aa$, and most likely $aa$ if they have many offspring that are all heterozygote.    


## General correlation matrices

